---
title: 'Project 2: Data Preparation & Transformation Pt. I'
author: "Albert Gilharry"
date: "06 March 2018"
output:
  html_document:
    css: ./css.css
    highlight: pygments
    pdf_document: default
    theme: cerulean
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Set I - US Chronic Disease Indicators (CDI)

<div id = "comment">
This data set was posted by **Niteen Kumar**. "The dataset actually outlines how different states are impacted by certain types of disease category along with clear indicators such as alcohol use among youth, Binge drinking prevalence among adults aged ≥ 18 years, Heavy drinking among adults aged ≥ 18 years, Chronic liver disease mortality etc." 

My goals with this data set are as follows: <br />

  (-) **Load, Tidy, and transform the data set.** <br />
  (-) **Investigate how each state and location are impacted by several types of chronic diseases.** <br />
</div>


## Load Libraries

```{r load-libraries}
library("tidyverse")
library("stringr")
library("DT")
library("leaflet")
library("geojsonio")
library("ggplot2")
library("fiftystater")
library("colorplaner")
library("mapproj")
```

## Load Data
```{r load-data}
data <- read.csv("data/U.S._Chronic_Disease_Indicators__CDI.csv", sep = ",", header = TRUE, stringsAsFactors=FALSE,fileEncoding="UTF-8-BOM")
```

## Preview Data
<div id = "solution">
```{r preview-data}
head(data)
```

```{r cols-data}
ncol(data)
```

```{r rows-data}
nrow(data)
```

</div>

<div id = "comment">
As we can see, this dataset is a very complex dataset, it has 19 columns, 53469 columns, and missing values.
Further more, the variables aren't standardized. For example, the `Year` variable may contains a single year value, or a range of years. 
The quantities in `DataValue` may be have different values, it may be per 100,000, nominal, etc. 

I will focus on the prevelence of alcohol related idicators accross different states.
</div>

(1) Select only cases with a category of Alchohol

<div id = "solution">

```{r filter-cancer}
data <- filter(data,Category == "Alcohol")
head(data)
```
</div>


(2) Remove columns that are not needed for analysis.

<div id = "solution">
```{r select-cols}
data <- select(data, Year, LocationDesc, Category, Indicator,IndicatorID, DataValueUnit, DataValueType, DataValue, GeoLocation)
datatable(data, options = list(filter = FALSE),filter="top")
```
</div>

  (3) Create columns to extract latitude and longitude
  
  <div id = "comment">
  Geospatial analysis typically requires separate structures for latitude and longitude. 
  We need to split the `GeoLocation` variable and remove additional punctuations to facilitate this.
  An example of a case of this variable is "(32.84057112200048, -86.63186076199969)"
  </div>
  
<div id = "solution">
```{r set-geo}
data$lats <- str_extract_all( data$GeoLocation, "-?[[:digit:]]*\\.[[:digit:]]*" )[[1]]
data$lons <- str_extract_all( data$GeoLocation, "-?[[:digit:]]*\\.[[:digit:]]*" )[[2]]
data <- select(data, -GeoLocation)
```
</div>


<div id = "solution">
Preview coordinates
```{r view-geo}
datatable(data, options = list(filter = FALSE),filter="top")
```
</div>

  (4) Map the prevlalence of binge drinking among adults accross states.
  
<div id = "comment">

There are multiple cases for the same `LocationDesc`,`Indicator`, and `DataValueType`. 
This would require grouping and summarizing before we can filter and map our indicator of interest.


I did not utilize the coordinates in this case. I used a package called `fifty_states`
to get the boundaries for each state. I filtered the data to meet the requirements of the query to produce the map.

This method can be used to plot most variables in dataset across states.
</div>

```{r map-indicator}

indicatorID <- 'ALC2_2'
indicator <- "Binge drinking prevalence among Adults Aged >= 18 years"
dataValueType <- 'Age-adjusted Prevalence'
data("fifty_states") # Load state properties

# perform grouping to condiser multiple entries for the same state
mapData <- data  %>%
  group_by(LocationDesc,IndicatorID,DataValueType) %>%
    summarise(DataValue = sum(DataValue) )
# consider only the indicator and values of interest
mapData <- filter( mapData, IndicatorID == indicatorID, DataValueType == dataValueType )

# the state data set is in lower case, create a column with lower case states for referencing
mapData <- data.frame( state = tolower(mapData$LocationDesc), mapData )

# plot map
p <- ggplot(mapData, aes(map_id = state)) + 
  # map points to the fifty_states shape data
  geom_map(aes(fill = DataValue), map = fifty_states, size = 0.15,color = "#ffffff") + 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  scale_fill_gradient2(name = dataValueType, midpoint = median(mapData$DataValue, na.rm = TRUE) ,
                       low = "blue", mid = "green",
                       high = "red") +
  coord_map() +
  labs(x = NULL, y = NULL,
       title = indicator) +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) 
  theme(legend.position = "bottom", 
        panel.background = element_blank())

p + fifty_states_inset_boxes() 
```

<div id = "comment"">
The resulting map shows that binge drinking among adults is more prominent in the northern states. 
North Dakota is the most prominent, could it be the cold weather blues?


This data set is very complex and I have only scratched the surface of its potential. 
The procedures above can be replicated to map and compare other indicators.


This data set provides an excellent learning opportunity.
</div>
